\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{enumitem}
\title{Phase 1 Technical Report: Dataset-Aware Relational Intelligence Extraction}
\author{RFM v2 Phase 1}
\date{\today}

\begin{document}
\maketitle

\section{Purpose}
Phase 1 builds reusable, dataset-aware artifacts for downstream task compilers. It is task-agnostic and enforces a global temporal cutoff derived from dataset validation starts.

\section{Pipeline Outputs}
For dataset $D$, artifacts are emitted under \texttt{<output-dir>/<D>/}:
\begin{itemize}[leftmargin=1.5em]
  \item \texttt{schema.json}
  \item \texttt{stats.json}
  \item \texttt{semantics.json}
  \item \texttt{semantic\_context\_graph.json}
  \item \texttt{path\_catalog.json}
  \item \texttt{safety\_rules.json}
\end{itemize}

\section{Refactored Module Layout}
The codebase is split for readability and extension:
\begin{itemize}[leftmargin=1.5em]
  \item \texttt{phase1\_pipeline.py}: orchestration and artifact assembly.
  \item \texttt{phase1\_utils.py}: JSON serialization, statistical primitives, role enums.
  \item \texttt{phase1\_fk.py}: FK candidate inference utilities.
  \item \texttt{phase1\_semantic\_graph.py}: semantic context graph construction.
\end{itemize}

\section{Temporal Policy}
A single global cutoff (minimum \texttt{val\_start} across tasks in the dataset) is used. All Phase 1 profiling and relation inference uses rows strictly earlier than this cutoff.

\section{FK Inference Summary}
Inferred FK candidates use overlap and naming compatibility constraints:
\begin{itemize}[leftmargin=1.5em]
  \item child column must be ID-like and non-generic;
  \item overlap ratio thresholding (high/medium confidence);
  \item lexical compatibility between child-ID stem and parent identifiers;
  \item uniqueness guardrails to avoid pseudo-links.
\end{itemize}

\section{Semantic Context Graph}
The graph contains table nodes and directed edges:
\begin{itemize}[leftmargin=1.5em]
  \item Node fields include role, row count, trusted measure columns, degree/hub stats.
  \item Edge fields include direction, source (provided/inferred), overlap ratio,
  cardinality risk, temporal alignment, and composite edge weight.
\end{itemize}

\section{LLM Usage}
Optional LLM annotation is used for semantic roles and optional path scoring. Heuristic fallback remains available for robustness.

\section{Extensibility Points}
\begin{itemize}[leftmargin=1.5em]
  \item Replace FK confidence thresholds in \texttt{phase1\_fk.py}.
  \item Add new role-affinity or edge-weight terms in \texttt{phase1\_semantic\_graph.py}.
  \item Add new per-column statistics in \texttt{phase1\_utils.py} and wire into \texttt{stats.json}.
\end{itemize}

\section{Reproducibility}
Default execution:
\begin{verbatim}
python phase1/phase1_pipeline.py \
  --dataset rel-f1 \
  --output-dir phase1/artifacts
\end{verbatim}

\end{document}
